CrypTool Script\footnote{ Background reading for the CrypTool program}\\ Mathematics and Cryptography

### 2003-04-07 script/en/chapters/numbertheory.tex
cvs id
be anmerkungen

% \def\QM {{,\kern -0.9 pt ,}}
\setcounter{theorem}{0}
\setcounter{definition}{0}

@@

%\begin{quote}
%{\em Joanne K. Rowling\footnote{Joanne K. Rowling,~``Harry Potter and the Philosopher’s Stone'', Carlsen, (c) 
%1997, chapter ``Through the trapdoor'', p. 310.}:}\newline This isn't magic – it's logic – a puzzle. A 
%lot of the greatest wizards haven't got an ounce of logic.
%\end{quote}

@@

For more details on prime numbers (e.g. how ``Fermat’s Little Theorem'' can be used to test extremely large 
numbers to determine whether they are prime), please refer to the article on \hyperlink{Kapitel_2}{prime numbers} in this script.

@@

The known laws apply:
\begin{itemize}
\item[\bf 1.] {\bf Associative law:}\index{Associative law} \\ 
    $((a+b) + c) {\rm ~(mod~ } m) \equiv  (a + (b+c)) {\rm ~(mod~ } m).$ \\
    $((a*b) * c) {\rm ~(mod~ } m) \equiv  (a * (b*c)) {\rm ~(mod~ } m).$
\item[\bf 2.] {\bf Commutative law:} \index{Commutative law}\\
    $(a+b) {\rm ~(mod~ } m) \equiv  (b+a) {\rm ~(mod~ } m).$ \\
     $(a*b) {\rm ~(mod~ } m) \equiv  (b*a) {\rm ~(mod~ } m).$
\end{itemize}
The associative law and commutative law apply to both addition and multiplication.
\begin{itemize}
\item[\bf 3.] {\bf Distributive law:} \index{Distributive law}\\
    $ (a * (b+c)) {\rm ~(mod~ } m) \equiv  (a*b + a*c) {\rm ~(mod~ } m).$ 
\item[\bf 4.] {\bf Reducibility:} \index{Reducibility} \\
    $(a+b) {\rm ~(mod~} m) \equiv  (a {\rm ~(mod~ } m) + b {\rm ~(mod~ } m)) {\rm ~(mod~} m).$ \\  
    $(a*b) {\rm ~(mod~} m) \equiv  (a {\rm ~(mod~ } m) * b {\rm ~(mod~ } m)) {\rm ~(mod~} m).$
\end{itemize}
The order in which order the modulo operation is performed is irrelevant.
\begin{itemize}
\item[\bf 5.] {\bf Existence of an identity (neutral element):} \index{Identity}\\
    $(a + 0) {\rm ~(mod~ } m) \equiv  (0 + a) {\rm ~(mod~ } m) \equiv  a {\rm ~(mod~ } m).$  \\
    $(a * 1) {\rm ~(mod~ } m) \equiv  (1 * a) {\rm ~(mod~ } m) \equiv  a {\rm ~(mod~ } m).$
\item[\bf 6.] {\bf Existence of an inverse element:} \\
    For all whole numbers $a$ and $m$ there exists a whole number $-a$ such that: \\
    $(a + (-a)) {\rm ~(mod~}m) \equiv  0 {\rm ~(mod~ } m)$ \quad (additive inverse).
\index{Inverse!additive}\\
    For each $a$ ($a \not\equiv 0 {\rm ~(mod~ } p$) ) where $p$ is prime there exists a whole number $a^{-1}$, 
such that: \\
    $(a * a^{-1}) {\rm ~(mod~ } p) \equiv 1 {\rm ~(mod~}p)$ \quad (multiplicative inverse). 
\index{Inverse!multiplicative}
\item[\bf 7.] \index{Closedness} {\bf Closedness}\footnote{\label{ftn-zth-closed}The property of closeness is always defined in relation to an operation in a set. See 
\hyperlink{Appendix_B}{Appendix B}.}:   \\
     $a, b \in G  \Longrightarrow  ( a + b ) \in G.$ \\
    $a, b \in G  \Longrightarrow  ( a * b ) \in G.$
\item[\bf 8.] \index{Transitivity} {\bf Transitivity}:

@@

Although it may seem difficult to calculate the table of multiplicative inverses for large moduli (this only applies 
to the areas of the table shaded dark grey), we can use Fermat's Little Theorem\index{Fermat!little theorem}
to create a simple algorithm for this \cite[p. 80]{Pfleeger1997}. Quicker algorithms are described, for instance, in 
\cite{Knuth1998}\footnote{Using Euclid’s extended theorem\index{Euclidean algorithm!extended}  (extended gcd), we can calculate the 
multiplicative inverse and determine whether numbers have an inverse (see \hyperlink{Appendix_A}{appendix 
A}). Alternatively, we can also use the primitive roots. }.

@@

\begin{itemize}
\item The set is closed \index{Closedness}
\item A neutral element exists
\item An inverse element exists
\item The associative law applies.
\end{itemize}

@@

We note that the definition of \emph{effort} is a crucial point: Bernstein
claims that effort is the product of time and costs of the machine
(including the memory used). The gist of the paper lies in the fact that he
can reduce a big part of factorizing to sorting. Using Schimmler's scheme,
sorting can be optimized by massive parallel computing.  At the end of
section 3 Bernstein explains this effect: The costs of $m^2$ parallel
computers with a constant amount of memory is a constant times $m^2$.  The
costs of a computer with a single processor and memory of size $m^2$ is
also of the order of $m^2$, but with a different constant factor.  With
$m^2$ processors in parallel, sorting of $m^2$ numbers (with Schimmler's
scheme) can be achieved in time $m$, while a $m^2$-memory computer needs
time of the order of $m^2$. Decreasing memory and increasing the number of
processors, the computing time can be reduced by a factor $1/m$ without
additional effort in terms of total costs.  In section 5 it is said that
massive parallel computing can also increase efficiency of factorizing
using Lenstra's elliptic-curve-method (a search algorithm has costs that
increase quadratically instead of cubically).

@@

The property of closeness \index{Closedness} is always defined in relation to an operation in a set.
The following shows how to construct the ``closed set'' $G$ with respect to the operation $+ {\rm ~(mod~} 8)$ for a given initial set $G_0$:
\begin{eqnarray*}
G_0 & = & \{ 2, 3 \} {\rm ~addition~of~the~numbers~in~} G_0 {\rm ~determines~further~numbers:} \nonumber \\
    & &    2 + 3 \equiv 5{\rm ~(mod~}8) = 5 \nonumber \\
    & &    2 + 2 \equiv 4{\rm ~(mod~}8) = 4 \nonumber \\
    & &    3 + 3 \equiv 6{\rm ~(mod~}8) = 6 \nonumber \\ 
G_1 & = & \{ 2, 3, 4, 5, 6 \} {\rm ~addition~of~the~numbers~in~} G_1 {\rm ~determines:}\nonumber \\
    & &    3 + 4 \equiv 7{\rm ~(mod~}8) = 7 \nonumber \\
    & &    3 + 5 \equiv 8{\rm ~(mod~}8) = 0 \nonumber \\
    & &    3 + 6 \equiv 9{\rm ~(mod~}8) = 1 \nonumber \\ 
G_2 & = & \{ 0, 1, 2, 3, 4, 5, 6, 7 \} {\rm ~~addition~of~the~numbers~in~} G_2 {~does~not~extend~the~set!} \nonumber \\
G_3 & = & G_2 {\rm ~~we~say:~} G_2 {\rm~is~closed~for~addition~~(mod~}8). \nonumber 
\end{eqnarray*}
End of forming a closed set.

>>>>

% $Id$
% \def\QM {{,\kern -0.9 pt ,}}
\setcounter{theorem}{0}
\setcounter{definition}{0}

@@

%\begin{quote}
%{\em Joanne K. Rowling\footnote{Joanne K. Rowling,~``Harry Potter and the Philosopher's Stone'', Carlsen, (c) 
%1997, chapter ``Through the trapdoor'', p. 310.}:}\newline This isn't magic – it's logic – a puzzle. A 
%lot of the greatest wizards haven't got an ounce of logic.
%\end{quote}

@@

For more details on prime numbers (e.g. how ``Fermat's Little Theorem'' can be used to test extremely large 
numbers to determine whether they are prime), please refer to the article on \hyperlink{Kapitel_2}{prime numbers} in this script.

@@

The known laws apply:
\begin{itemize}
\item[\bf 1.] {\bf Associative law:}\index{Associative law} \\ 
    $((a+b) + c) {\rm ~(mod~ } m) \equiv  (a + (b+c)) {\rm ~(mod~ } m).$ \\
    $((a*b) * c) {\rm ~(mod~ } m) \equiv  (a * (b*c)) {\rm ~(mod~ } m).$
\item[\bf 2.] {\bf Commutative law:} \index{Commutative law}\\
    $(a+b) {\rm ~(mod~ } m) \equiv  (b+a) {\rm ~(mod~ } m).$ \\
     $(a*b) {\rm ~(mod~ } m) \equiv  (b*a) {\rm ~(mod~ } m).$
\end{itemize}
The associative law and commutative law apply to both addition and multiplication.
\begin{itemize}
\item[\bf 3.] {\bf Distributive law:} \index{Distributive law}\\
    $ (a * (b+c)) {\rm ~(mod~ } m) \equiv  (a*b + a*c) {\rm ~(mod~ } m).$ 
\item[\bf 4.] {\bf Reducibility:} \index{Reducibility} \\
    $(a+b) {\rm ~(mod~} m) \equiv  (a {\rm ~(mod~ } m) + b {\rm ~(mod~ } m)) {\rm ~(mod~} m).$ \\  
    $(a*b) {\rm ~(mod~} m) \equiv  (a {\rm ~(mod~ } m) * b {\rm ~(mod~ } m)) {\rm ~(mod~} m).$
\end{itemize}
The order in which order the modulo operation is performed is irrelevant.
\begin{itemize}
\item[\bf 5.] {\bf Existence of an identity (neutral element):} \index{Identity}\\
    $(a + 0) {\rm ~(mod~ } m) \equiv  (0 + a) {\rm ~(mod~ } m) \equiv  a {\rm ~(mod~ } m).$  \\
    $(a * 1) {\rm ~(mod~ } m) \equiv  (1 * a) {\rm ~(mod~ } m) \equiv  a {\rm ~(mod~ } m).$
\item[\bf 6.] {\bf Existence of an inverse element:} \\
    For all whole numbers $a$ and $m$ there exists a whole number $-a$ such that: \\
    $(a + (-a)) {\rm ~(mod~}m) \equiv  0 {\rm ~(mod~ } m)$ \quad (additive inverse).
\index{Inverse!additive}\\
    For each $a$ ($a \not\equiv 0 {\rm ~(mod~ } p$) ) where $p$ is prime there exists a whole number $a^{-1}$, 
such that: \\
    $(a * a^{-1}) {\rm ~(mod~ } p) \equiv 1 {\rm ~(mod~}p)$ \quad (multiplicative inverse). 
\index{Inverse!multiplicative}
\item[\bf 7.] \index{Closeness} {\bf Closeness}\footnote{\label{ftn-zth-closed}The property of closeness is always defined in relation to an operation in a set. See 
\hyperlink{Appendix_B}{Appendix B}.}:   \\
     $a, b \in G  \Longrightarrow  ( a + b ) \in G.$ \\
    $a, b \in G  \Longrightarrow  ( a * b ) \in G.$
\item[\bf 8.] \index{Transitivity} {\bf Transitivity}:

@@

Although it may seem difficult to calculate the table of multiplicative inverses for large moduli (this only applies 
to the areas of the table shaded dark grey), we can use Fermat's Little Theorem\index{Fermat!little theorem}
to create a simple algorithm for this \cite[p. 80]{Pfleeger1997}. Quicker algorithms are described, for instance, in 
\cite{Knuth1998}\footnote{Using Euclid's extended theorem\index{Euclidean algorithm!extended}  (extended gcd), we can calculate the 
multiplicative inverse and determine whether numbers have an inverse (see \hyperlink{Appendix_A}{appendix 
A}). Alternatively, we can also use the primitive roots. }.

@@

\begin{itemize}
\item The set is closed \index{Closeness}
\item A neutral element exists
\item An inverse element exists
\item The associative law applies.
\end{itemize}

@@

We note that the definition of \emph{effort} is a crucial point: Bernstein
claims that effort is the product of time and costs of the machine
(including the memory used). The gist of the paper lies in the fact that he
can reduce a big part of factorizing to sorting. Using Schimmler's scheme,
sorting can be optimized by massive parallel computing.  At the end of
section 3 Bernstein explains this effect: The costs of $m^2$ parallel
computers with a constant amount of memory is a constant times $m^2$.  The
costs of a computer with a single processor and memory of size $m^2$ is
also of the order of $m^2$, but with a different constant factor.  With
$m^2$ processors in parallel, sorting of $m^2$ numbers (with Schimmler's
scheme) can be achieved in time $m$, while a $m^2$-memory computer needs
time of the order of $m^2$. Decreasing memory and increasing the number of
processors, the computing time can be reduced by a factor $1/m$ without
additional effort in terms of total costs.  In section 5 it is said that
massive parallel computing can also increase efficiency of factorizing
using Lenstra's elliptic-curve-method (a search algorithm has costs that
increase in a quadratic square manner instead of cubically).

@@

The property of closeness \index{Closeness} is always defined in relation to an operation in a set.
The following shows how to construct the ``closed set'' $G$ with respect to the operation $+ {\rm ~(mod~} 8)$ for a given initial set $G_0$:
\begin{eqnarray*}
G_0 & = & \{ 2, 3 \} {\rm ~addition~of~the~numbers~in~} G_0 {\rm ~determines~further~numbers:} \nonumber \\
    & &    2 + 3 \equiv 5{\rm ~(mod~}8) = 5 \nonumber \\
    & &    2 + 2 \equiv 4{\rm ~(mod~}8) = 4 \nonumber \\
    & &    3 + 3 \equiv 6{\rm ~(mod~}8) = 6 \nonumber \\ 
G_1 & = & \{ 2, 3, 4, 5, 6 \} {\rm ~addition~of~the~numbers~in~} G_1 {\rm ~determines:}\nonumber \\
    & &    3 + 4 \equiv 7{\rm ~(mod~}8) = 7 \nonumber \\
    & &    3 + 5 \equiv 8{\rm ~(mod~}8) = 0 \nonumber \\
    & &    3 + 6 \equiv 9{\rm ~(mod~}8) = 1 \nonumber \\ 
G_2 & = & \{ 0, 1, 2, 3, 4, 5, 6, 7 \} {\rm ~~addition~of~the~numbers~in~} G_2 {~does~not~extend~the~set!} \nonumber \\
G_3 & = & G_2 {\rm ~~we~say:~} G_2 {\rm~is~closed~for~addition~~(mod~}8). \nonumber 
\end{eqnarray*}
End of forming a closed set.

### 2003-04-07 script/en/chapters/primes.tex
cvs id
be anmerkungen

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% C H A P T E R   T W O
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

Although there are only $92$ natural chemical elements, the number of prime
numbers is unlimited. Even the Greek, \index{Euclid} Euclid%
\footnote{Euclid,
a Greek mathematician of 4th and 3rd century B.C. He worked at the
Egyptian academy of Alexandria and wrote ``The Elements'', the most well 
known systematically textbook of the Greek mathematics.} 
knew this in the third century B.C.
\begin{theorem}[Euclid\footnote{The common usage of the term does not denote Euclid as the inventor of the theorem rather;
the true inventor is merely not as prominent. The theorem has already been distinguished
and proven in Euclid’s Elements (Book IX, s. 20). The phraseology is remarkable due to 
the fact that the word infinite is not used. The text reads as followed
$$
O\acute{\iota}~\pi\varrho\tilde{\omega}\tau o \iota~\grave{\alpha}\varrho\iota\vartheta\mu o\grave{\iota}~
\pi\lambda\varepsilon\acute{\iota}o \upsilon\varsigma~\varepsilon\grave{\iota}\sigma\grave{\iota}~
\pi\alpha\nu\tau\grave{o}\varsigma~\tau o \tilde{\upsilon}~
\pi\varrho o \tau\varepsilon\vartheta\acute{\varepsilon}\nu\tau o \varsigma~
\pi\lambda\acute{\eta}\vartheta\ o \upsilon\varsigma~
\pi\varrho\acute{\omega}\tau\omega\nu~
\grave{\alpha}\varrho\iota\vartheta\mu\tilde{\omega}\nu,
$$
the English translation of which is: the prime numbers are more than 
any previously existing amount of prime numbers.
}]\label{thm-pz-euklid} % Ende der Fußnote
% Fussnote VOR "]" in [Euclid], damit kein Leerraum vor der Fussnotennummer.
% Vorher stand da: ...(Euclid). BLANK Fussnote und das Blank stoerte.
% Nun steht die Fussnote direkt hinter "Euclid" und vor der ")".
% Eigentlich hätte ich sie gerne direkt hinter "(Euclid)", noch vor dem 
% automatisch gesetzten Punkt. 
The sequence of prime numbers does not discontinue.
Therefore, the quantity of prime numbers is infinite.
\end{theorem}
His proof that there is an infinite number of
primes is still considered to be a brilliant mathematical consideration and
conclusion today (proof by contradiction). He assumed that there is only a
finite number of primes and therefore a largest prime number. Based on this
assumption, he drew logical conclusions until he obtained an obvious
contradiction. This meant that something must be wrong. As there were no
mistakes in the chain of conclusions, it could only be the assumption that was
wrong. Therefore, there must be an infinite number of primes!

@@

\vskip + 5pt
\hypertarget{Mer-nums-not-always-prim}{}
Unfortunately this theorem only applies in one direction (the inverse 
statement does not apply, no equivalence): that means that there exist 
prime exponent for which the Mersenne number ist {\bf not} prime (see the 
above example $2^{11}-1, $ where $11$ is prime, but $2^{11}-1$ not).

@@

As early as in the 19th century, it was discovered that the $29$-digit number $$
F(7) = 2^{2^7} + 1 $$ is not prime. However, it was not until 1970 that
Morrison/Billhart managed to factorise it.
\begin{eqnarray*}
F(7) & = & 34,279,974,696,877,740,253,374,607,431,768,211,457 \\
& = & 59, 649, 589, 127, 497, 217 \cdot  574, 689, 200, 685,
129,054, 721.
\end{eqnarray*}

>>>>

% $Id$
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% C H A P T E R   T W O
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

Although there are only $92$ natural chemical elements, the number of prime
numbers is unlimited. Even the Greek, \index{Euclid} Euclid%
\footnote{Euclid,
a Greek mathematician of 4th and 3rd century B.C. He worked at the
Egyptian academy of Alexandria and wrote ``The Elements'', the most well 
known systematically textbook of the Greek mathematics.} 
knew this in the third century B.C.
\begin{theorem}[Euclid\footnote{The common usage of the term does not denote Euclid as the inventor of the theorem rather;
the true inventor is merely not as prominent. The theorem has already been distinguished
and proven in Euclid's Elements (Book IX, s. 20). The phraseology is remarkable due to 
the fact that the word infinite is not used. The text reads as followed
$$
O\acute{\iota}~\pi\varrho\tilde{\omega}\tau o \iota~\grave{\alpha}\varrho\iota\vartheta\mu o\grave{\iota}~
\pi\lambda\varepsilon\acute{\iota}o \upsilon\varsigma~\varepsilon\grave{\iota}\sigma\grave{\iota}~
\pi\alpha\nu\tau\grave{o}\varsigma~\tau o \tilde{\upsilon}~
\pi\varrho o \tau\varepsilon\vartheta\acute{\varepsilon}\nu\tau o \varsigma~
\pi\lambda\acute{\eta}\vartheta\ o \upsilon\varsigma~
\pi\varrho\acute{\omega}\tau\omega\nu~
\grave{\alpha}\varrho\iota\vartheta\mu\tilde{\omega}\nu,
$$
the English translation of which is: the prime numbers are more than 
any previously existing amount of prime numbers.
}]\label{thm-pz-euklid} % Ende der Fußnote
% Fussnote VOR "]" in [Euclid], damit kein Leerraum vor der Fussnotennummer.
% Vorher stand da: ...(Euclid). BLANK Fussnote und das Blank stoerte.
% Nun steht die Fussnote direkt hinter "Euclid" und vor der ")".
% Eigentlich hätte ich sie gerne direkt hinter "(Euclid)", noch vor dem 
% automatisch gesetzten Punkt. 
The sequence of prime numbers does not discontinue.
Therefore, the quantity of prime numbers is infinite.
\end{theorem}
His proof that there is an infinite number of
primes is still considered to be a brilliant mathematical consideration and
conclusion today (proof by contradiction). He assumed that there is only a
finite number of primes and therefore a largest prime number. Based on this
assumption, he drew logical conclusions until he obtained an obvious
contradiction. This meant that something must be wrong. As there were no
mistakes in the chain of conclusions, it could only be the assumption that was
wrong. Therefore, there must be an infinite number of primes!

@@

\vskip + 5pt
\hypertarget{Mer-nums-not-always-prim}{}
Unfortunately this theorem only applies in one direction (the inverse 
statement does not apply, no equivalence): that means that there exist 
prime exponent for which the Mersenne number is {\bf not} prime (see the 
above example $2^{11}-1, $ where $11$ is prime, but $2^{11}-1$ not).

@@

As early as in the 19th century, it was discovered that the $29$-digit number $$
F(7) = 2^{2^7} + 1 $$ is not prime. However, it was not until 1970 that
Morrison/Billhart managed to factorise it.
\begin{eqnarray*}
F(7) & = & 340,282,366,920,938,463,463,374,607,431,768,211,457 \\
& = & 59, 649, 589, 127, 497, 217 \cdot  5,704,689,200,685,129,054,721
\end{eqnarray*}
